The host galaxy of the short GRB~111117A at $z = 2.211$\thanks{Based on observations collected at ESO/VLT under programme 088.A-0051 and 091.D-0904, at TNG under programme A24TAC\_38, at Gemini North under programme GN-2011B-Q-10 and GTC under programme GTC43-11B.}

### 2017-06-29 GRB111117A.tex
Final edits

\title{The host galaxy of the short GRB~111117A at $z = 2.211$\thanks{Based on observations collected at the European Southern Observatory, Paranal, Chile, Program ID: 088.A-0051 and 091.D-0904.}}

@@

\abstract{
It is notoriously difficult to localize short $\gamma$-ray bursts (sGRBs) and
their hosts to measure their redshifts. These measurements, however, are
critical to constrain the nature of sGRB progenitors, the redshift distribution
of these events and the $r$-process element enrichment history of
the universe. Here, we present spectroscopy of the host galaxy of GRB~111117A
and measure its redshift to be $z = 2.211$. This makes GRB~111117A the most
distant high-confidence short duration GRB detected to date. Our spectroscopic
redshift supersedes a lower, previously estimated photometric redshift value for
this burst.

@@

\keywords{Gamma-ray burst: individual: GRB~111117A -- Gamma-ray burst: general -- Galaxies: high-redshift -- binaries: general -- X-rays: bursts -- Techniques: imaging spectroscopy}

@@

The \textit{Swift} satellite \citep{Gehrels2004} greatly improved the
understanding of sGRB progenitors thanks to its quick localization capability.
The bulk of these localizations have associated galaxies at relatively low
redshifts with a median redshift of $z\sim0.5$ \citep{Berger2014}, and because
most of these measurements come from the associated hosts, it is arguably biased
towards lower redshifts.

The host galaxies of sGRBs are diverse. They are more massive and less actively
star-forming on average than lGRB hosts \citep{Fong2013b}, while in some cases,
no host galaxy can be identified above detection threshold of deep follow-up
observations \citep{Berger2010a, Tunnicliffe2014}. Together with their position
within their hosts \citep{Fong2013a}, this suggests a progenitor system that can
be very long lived in comparison to lGRBs, and is tracing with host stellar mass rather than
star-formation rate (SFR).

The electromagnetic signals from sGRBs are a promising channel to accurately
localize NS mergers, which holds the promise for a detection of an associated
gravitational wave (GW) signal \citep{Ghirlanda2016}. The simultaneous detection
of the sGRB and the GW will provide new a promising ways to constrain the binary
inclination angle \citep{Arun2014} and measure cosmological distances
\citep{Nissanke2010}.

@@

Spectroscopic observations were carried out using the cross-dispersed echelle
spectrograph, VLT/X-shooter \citep{Vernet2011}, at four seperate epochs. The
burst was observed 38 hours after the Burst Alert Telescope (BAT) trigger under
ESO programme 088.A-0051 (PI: Fynbo) and again later under ESO programme
091.D-0904 (PI: Hjorth). Observations use a simple ABBA nodding pattern, using 5
arcsec nod throws. X-shooter covers the wavelength range from 3000~\AA~to
24\,800~\AA~(21\,000~\AA~when the $K$-band blocking filter is used) across three
spectroscopic arms. We carried out the bias-correction, flat-fielding, order
tracing, wavelength calibration, rectification, and flux calibration using the
VLT/X-shooter pipeline version 2.8.4 \citep{Modigliani2010} run in physical
mode. Because the echelle orders are curved across each detector, a rectification
algorithm is employed which introduces correlations between neighboring pixels.
We select a pixel-scale of 0.2/0.2/0.6 \AA/pix for the UVB/VIS/NIR arm to
minimize the degree of correlation while conserving the maximal resolution. The
observations are combined and extracted using scripts described in Selsing et
al. 2017 (in prep.) and available
online\footnote{\url{https://github.com/jselsing/XSGRB_reduction_scripts}}. The
signal-to-noise ratio of the continuum in the near-infrared arm is too low to
use the optimal extraction algorithm \citep{Horne1986} and therefore the
extraction is carried out with a simple aperture covering the entire host. An
overview of the spectroscopic observations is given in
Table~\ref{tab:spec_overview}, and the slit position is shown in
Fig.~\ref{fig:spec_setup}. We show the extracted spectrum in Fig.~\ref{fig:SED}.

@@

Using the luminosity of \ha, we can infer the star-formation rate (SFR) of a
galaxy \citep{Kennicutt1998}. At the redshift of the GRB host, \ha~is observed
at 21\,000~\AA~where the night sky is very bright. In addition, several bright
sky-lines are superposed on the line, making an accurate estimate of the
\ha-flux difficult. We obtain a limit on the SFR by numerically integrating the
part of \ha~free of contamination and correcting for the missing flux using the
line profile and obtain $F_{H \alpha} > 4.1 \times 10^{-17} \mathrm{erg}
\mathrm{s}^{-1} \mathrm{cm}^{-2}$. After converting the \citet{Kennicutt1998}
relation to a \citet{Chabrier2003} initial mass function using
\citet{Madau2014}, we derive a limit of $SFR > 7 M_\odot$~yr$^{-1}$.
Additionally, based on the width of \oiii, an aperture covering the line is
integrated over, where synthetic sky lines \citep{Noll2012, Jones2013} have been
masked and interpolated over. From the integrated \ha-line, we estimate SFR =
$18 \pm 3$ M$_\odot$ yr$^{-1}$. From the SED-fit (Sect.~\ref{SED}), and the
detection of \lya, the host is constrained to contain very little or no dust,
although the presence of Lya does not exclude dust. Therefore we do not apply a
dust-correction to the measured \ha~flux here. \oii is close to a region of
strong telluric absorption, which is why no SFR is inferred from this line.

@@

A measure of the optical-to-X-ray flux ratio is parametrized in terms of the
"darkness"-parameter $\beta_{OX} $\citep{Jakobsson2004}. Using the optical
afterglow limits\citep{Cucchiara2011, Cenko2011}, the X-ray lightcurve can be
interpolated and evaluated at the time of the non-detection. We find $\beta_{OX}
< 0.79$, consistent with what was reported in \citet{Sakamoto2013}.

@@

As pointed out by \citet{Margutti2012} and \citet{Sakamoto2013}, GRB~111117A is
securely classified as a sGRBs. Because the observed classification indicators,
$T_{90}$ and hardness ratio, do not depend strongly on redshift
\citep{Littlejohns2013a}, the updated redshift does not change this designation.
The intrinsic spectral lag shortens, but since it is already consistent with
zero, this does not affect the classification.

@@

\citet{Bromberg2013} investigated the degree to which the long and short
population distributions overlap and quantified the certainty in the class
membership. According to \citet{Bromberg2013}, GRB~111117A has $96_{-5}^{+3}$
percent probability of being a sGRB. Compared to the other two sGRB candidates
at high redshift, GRB~060121 \citep{DeUgartePostigo2006, Levan2006} at $1.7
\lesssim z \lesssim 4.5$ ($17_{-15}^{+14}$ per cent) and GRB~090426
\citep{Antonelli2009, Levesque2010, Thone2011} at $z = 2.609$ ($10_{-10}^{+15}$
per cent), the certainty in class membership for GRB~111117A is much higher.

@@

The redshift and secure classification of GRB~111117A means that it occurred
when the universe was younger by 1.8 Gyr compared to any other non-collapsar GRB
ever detected. This number is 3.2 Gyr for the next-highest spectroscopic
redshift. If the merger of NSs is the primary agent for the $r$-process element
enrichment of the universe \citep{Goriely2011, Ji2016, Komiya2016}, GRB~111117A
marks the earliest detection of this process.

@@

We show the recalculated $N_\mathrm{H}$ in Fig.~\ref{fig:NH_z} where we compare
with the distributions of complete samples of both long and short GRBs. The lGRB
sample is from \citet{Arcodia2016} and the sGRB sample is from
\citet{DAvanzo2014a}. From the sGRB sample of \citet{DAvanzo2014a} we have
excluded GRB~090426 which does not belong in a short sample as highlighted in
Sect. \ref{classification}. Both comparison samples consider the largest
temporal interval of constant hardness ratio to prevent spectral changes that
can affect the X-ray derived column density. The 17 (5) of the 99 (15) long
(short) GRBs which do not have redshift have been excluded from our analysis.

@@

From the clear host association, GRB~111117A does not belong to the hostless
class of GRBs \citep{Berger2010a} and because the host exhibits emission lines
this is indicative of a population of relatively young stars. As the majority of
sGRBs \citep{Fong2013b}, the host of GRB~111117A is a late-type galaxy and is
entirely consistent in terms of stellar mass and stellar age with the general
population of sGRB hosts ($\left\langle M _* \right\rangle = 10^{10.1}
M_{\odot}$ and $\left\langle \tau _* \right\rangle = 0.3 $~Gyr)
\citep{Leibler2010}. Being a late-type host, both the stellar mass and sSFR are
entirely within the range expected for the hosts of sGRBs \citep{Behroozi2014}.

@@

A single sGRB at high redshift does little in terms of constraining the redshift
distribution of sGRBs. In particular, other sGRB hosts could be missed because
they are intrinsically fainter and thus this high-$z$ event is only detected due
to the brightness of it's host. \citet{Berger2014} compiled a sample of sGRB
host luminosities, normalized by the characteristic galaxy luminosity at their
respective redshift, $L_B/L^{\star}_{B}$. To convert the SED-inferred $M_B$ of
GRB~111117A to $L_B/L^{\star}_{B}$, we use the characteristic absolute $B$-band
magnitude of the Schechter function for blue galaxies ($U - V < 0.25$) in the
redshift window $2.0 \leq z \leq 2.5$ from \citet{Marchesini2007} and obtain
$L_B/L^{\star}_{B} = 1.2$.

Using the complete, flux-limited selection of bursts from \citet{DAvanzo2014a},
excluding GRB~111117A and the non-sGRB GRB~090426, we have a statistically
homogeneous sample from which we can address the implications of the redshift of
GRB~111117A. Out of the 14 hosts in the sample, 10 (71 per cent) have measured
redshifts and $L_B/L^{\star}_{B}$. Of the complete sample, the host of
GRB~111117A  is brighter than 80 per cent of the hosts with measured
$L_B/L^{\star}_{B}$. If we conservatively assume that \textit{all} the host
missing $L_B/L^{\star}_{B}$ are brighter than the host of GRB~111117A, the host
still brighter than > 60 per cent of sGRB hosts. For all 26 hosts with
$L_B/L^{\star}_{B}$ from \citet{Berger2014}, the host of GRB~111117A is brighter
than 73 per cent.

If we assume that we are able to obtain emission-line redshifts from hosts with
$R < 24.5$~mag \citep{Kruhler2012}, then we would have missed 60 per cent
of the redshifts (6 out of 10 hosts), if they were at the redshift of
GRB~111117A due to the hosts being too faint. The corresponding number is around
45 per cent (12 out of 26) from the full sample of \citet{Berger2014}, reflecting
the lower mean $L_B/L^{\star}_{B}$ of the complete sample. Because the average
SFR of galaxies hosting lGRBs is higher than for galaxies hosting sGRBs, the
fraction of missed burst redshifts is likely higher although the cosmic SFR
evolution could play a role in improving redshift determinability.

A fraction of the bursts missing redshift are host-less and is therefore likely
at moderate redshifts \citep{Tunnicliffe2014}, but should some of the remainder
be at high redshift, the missed fraction will increase. If we assume that
\textit{all} the bursts that are missing redshifts are at high-$z$ and missed
due to host faintness, 10 out of 14 hosts in the complete sample (71 per cent)
would be missed at $z = 2.211$. This serves as an upper limit on the fraction of
missed burst redshifts at high-$z$. Conversely, if all bursts missing redshift
are at low redshift and missed for other reasons, 6 out of 14 hosts (43 per
cent) would be missed at $z = 2.211$. The two limits indicate that we are
missing between 43 and 71 per cent of sGRB hosts at $z sim 2$ due to host
faintness. For the 13 hosts missing redshift in \citet{Berger2014}, this means
that between 6 and 9 of these should be at high redshift, but missed due to host
faintness.

@@

Using the new distance, the X-ray derived $n_H$ towards GRB~111117A is the
highest within a complete sample of sGRB hosts and is consistent with the
$n_H-z$ evolution traced by the hosts of lGRBs. The SFR of the host is in the
upper end of the sGRB host SFR distribution and no dust is present. The high
$N_H$ is difficult to reconcile with the large projected host offset and the
absence of dust. One possible explanation could be, that GRB~111117A is formed
through a prompt channel of sGRB formation and originates in a star forming
region located in the outskirts of the host.

@@

Using the age of the universe at the time of explosion allows us to set
constraints on the maximal separation between the engine constituents at the
time of formation. We find that the maximal separation of two NSs at
formation time is $a_0 < 3.2~R_\odot$, which excludes some of the formation
channels with the longest timescales.

@@

\begin{acknowledgements}

TK acknowledges support through the Sofja Kovalevskaja Award to P. Schady.
% 
JJ acknowledges support from NOVA and NWO-FAPESP grant for advanced
instrumentation in astronomy.
%
CT acknowledges support from a Spanish National Research Grant of Excellence
under project AYA 2014-58381-P and funding associated to a Ramón y Cajál
fellowship under grant number RyC-2012-09984.
%
AdUP acknowledges support from a Ramón y Cajal fellowship, a BBVA Foundation
Grant for Researchers and Cultural Creators,and the Spanish Ministry of Economy
and Competitiveness through project AYA2014-58381-P. Partly based on
observations made with the Gran Telescopio Canarias (GTC).
%
ZC acknowledges support from the Spanish research project AYA 2014-58381-P and
support from Juan de la Cierva Incorporaci\'on fellowships IJCI-2014-21669.
%
RSR acknowledges AdUP's BBVA Foundation Grant for Researchers and Cultural
Creators and support from ASI (Italian Space Agency) through the Contract n. 2015-046-R.0 and from European Union Horizon 2020 Programme under the AHEAD project (grant agreement n. 654215).
%
This research made use of Astropy, a community-developed core Python package for Astronomy \citep{TheAstropyCollaboration2013}. The analysis and plotting has been achieved using the Python-based packages Matplotlib \citep{Hunter2007}, Numpy and Scipy \citep{VanderWalt2011} along with other community-developed packages.

>>>>

\title{The host galaxy of the short GRB~111117A at $z = 2.211$\thanks{Based on observations collected at ESO/VLT under programme 088.A-0051 and 091.D-0904, at TNG under programme A24TAC\_38, at Gemini North under programme GN-2011B-Q-10 and GTC under programme GTC43-11B.}}

@@

\abstract{
It is notoriously difficult to localize short $\gamma$-ray bursts (sGRBs) and
their hosts to measure their redshifts. These measurements, however, are
critical to constrain the nature of sGRB progenitors, their redshift
distribution and the $r$-process element enrichment history of the universe.
Here, we present spectroscopy of the host galaxy of GRB~111117A and measure its
redshift to be $z = 2.211$. This makes GRB~111117A the most distant
high-confidence short duration GRB detected to date. Our spectroscopic redshift
supersedes a lower, previously estimated photometric redshift value for this
burst.

@@

\keywords{gamma-ray burst: individual: GRB~111117A -- gamma-ray burst: general -- galaxies: high-redshift -- binaries: general -- X-rays: bursts -- techniques: imaging spectroscopy}

@@

The \textit{Swift} satellite \citep{Gehrels2004} greatly improved the
understanding of sGRB progenitors thanks to its quick localization capability.
The bulk of these localizations have associated galaxies at relatively low
redshifts with a median redshift of $z\sim0.5$ \citep{Berger2014}, and because
most of these measurements come from the associated hosts, it is arguably biased
towards lower redshifts. The host galaxies of sGRBs are diverse. They are more
massive and less actively star-forming on average than lGRB hosts
\citep{Fong2013b}, while in some cases, no host galaxy can be identified above
detection threshold of deep follow-up observations \citep{Berger2010a,
	Tunnicliffe2014}. Together with their position within their hosts
\citep{Fong2013a}, this suggests a progenitor system that can be very long lived
in comparison to lGRBs, and is tracing with host stellar mass rather than
star-formation rate (SFR).

The electromagnetic signals from sGRBs are a promising channel to accurately
localize NS mergers, which holds the promise for a detection of an associated
gravitational wave (GW) signal \citep{Ghirlanda2016}. The simultaneous detection
of a sGRB and a GW will provide new a promising ways to constrain the binary
inclination angle \citep{Arun2014} and measure cosmological distances
\citep{Nissanke2010}.

@@

Spectroscopic observations were carried out using the cross-dispersed echelle
spectrograph, VLT/X-shooter \citep{Vernet2011}, at four seperate epochs. The
burst was observed 38 hours after the Burst Alert Telescope (BAT) trigger under
ESO programme 088.A-0051 (PI: Fynbo) and again later under ESO programme
091.D-0904 (PI: Hjorth). Observations use a simple ABBA nodding pattern, using 5
arcsec nod throws. X-shooter covers the wavelength range from 3000~\AA~to
24\,800~\AA~(21\,000~\AA~when the $K$-band blocking filter is used) across three
spectroscopic arms. We carried out the bias-correction, flat-fielding, order
tracing, wavelength calibration, rectification, and flux calibration using the
VLT/X-shooter pipeline version 2.8.4 \citep{Modigliani2010} run in physical
mode. Because the echelle orders are curved across each detector, a
rectification algorithm is employed which introduces correlations between
neighboring pixels. We select a pixel-scale of 0.2/0.2/0.6 \AA/pix for the
UVB/VIS/NIR arm to minimize the degree of correlation while conserving the
maximal resolution. The observations are combined and extracted using scripts
described in Selsing et al. 2017 (in prep.) and available
online\footnote{\url{https://github.com/jselsing/XSGRB_reduction_scripts}}. The
signal-to-noise ratio of the continuum in the near-infrared arm is too low to
use the optimal extraction algorithm \citep{Horne1986} and therefore the
extraction is carried out with a simple aperture covering the entire host, where
we sum the flux in the aperture and derive the error by summing the
corresponding pixel errors in quadrature. An overview of the spectroscopic
observations is given in Table~\ref{tab:spec_overview}, and the slit position is
shown in Fig.~\ref{fig:spec_setup}. We show the extracted spectrum in
Fig.~\ref{fig:SED}.

@@

Using the luminosity of \ha, we can infer the star-formation rate (SFR) of the
host \citep{Kennicutt1998}. At the redshift of the GRB host, \ha~is observed
at 21\,000~\AA~where the night sky is very bright. In addition, several bright
sky-lines are superposed on the line, making an accurate estimate of the
\ha-flux difficult. We obtain a limit on the SFR by numerically integrating the
part of \ha~free of contamination and correcting for the missing flux using the
line profile and obtain $F_{H \alpha} > 4.1 \times 10^{-17} \mathrm{erg}
\mathrm{s}^{-1} \mathrm{cm}^{-2}$. After converting the \citet{Kennicutt1998}
relation to a \citet{Chabrier2003} initial mass function using
\citet{Madau2014}, we derive a limit of $SFR > 7 M_\odot$~yr$^{-1}$.
Additionally, based on the width of \oiii, an aperture covering the line is
integrated over, where synthetic sky lines \citep{Noll2012, Jones2013} have been
masked and interpolated over. From the integrated \ha-line, we estimate SFR =
$18 \pm 3$ M$_\odot$ yr$^{-1}$. From the SED-fit (Sect.~\ref{SED}), and the
detection of \lya, the host is constrained to contain very little or no dust,
although the presence of Lya does not exclude dust. Therefore we do not apply a
dust-correction to the measured \ha~flux here. \oii~is close to a region of
strong telluric absorption, which is why no SFR is inferred from this line.

@@

A measure of the optical-to-X-ray flux ratio is parametrized in terms of the
"darkness"-parameter $\beta_{OX} $ \citep{Jakobsson2004}. Using the optical
afterglow limits \citep{Cucchiara2011, Cenko2011}, the X-ray lightcurve can be
interpolated and evaluated at the time of the non-detection. We find $\beta_{OX}
< 0.79$, consistent with what was reported in \citet{Sakamoto2013}.

@@

As pointed out by \citet{Margutti2012} and \citet{Sakamoto2013}, GRB~111117A is
securely classified as a sGRB. Because the observed classification indicators,
$T_{90}$ and hardness ratio, do not depend strongly on redshift
\citep{Littlejohns2013a}, the updated redshift does not change this designation.
The intrinsic spectral lag shortens, but since it is already consistent with
zero, this does not affect the classification.

@@

\citet{Bromberg2013} investigated the degree to which the long and short
population distributions overlap and quantified the certainty in class
membership. According to \citet{Bromberg2013}, GRB~111117A has $96_{-5}^{+3}$
percent probability of being a sGRB. Compared to the other two sGRB candidates
at high redshift, GRB~060121 \citep{DeUgartePostigo2006, Levan2006} at $1.7
\lesssim z \lesssim 4.5$ ($17_{-15}^{+14}$ per cent) and GRB~090426
\citep{Antonelli2009, Levesque2010, Thone2011} at $z = 2.609$ ($10_{-10}^{+15}$
per cent), the certainty in class membership for GRB~111117A is much higher.

@@

The redshift and secure classification of GRB~111117A mean that it occurred
when the universe was younger by 1.8 Gyr compared to any other non-collapsar GRB
ever detected. This number is 3.2 Gyr for the next-highest spectroscopic
redshift. If the merger of NSs is the primary agent for the $r$-process element
enrichment of the universe \citep{Goriely2011, Ji2016, Komiya2016}, this
marks the earliest detection of this process.

@@

We show the recalculated $N_\mathrm{H}$ in Fig.~\ref{fig:NH_z} where we compare
with the distributions of complete samples of both long and short GRBs. The lGRB
sample is from \citet{Arcodia2016} and the sGRB sample is from
\citet{DAvanzo2014a}. From the sGRB sample of \citet{DAvanzo2014a} we have
excluded GRB~090426 which does likely not belong in a short sample as highlighted in
Sect. \ref{classification}. Both comparison samples consider the largest
temporal interval of constant hardness ratio to prevent spectral changes that
can affect the X-ray derived column density. The 17 (5) of the 99 (15) long
(short) GRBs which do not have redshift have been excluded from our analysis.

@@

From the clear host association, GRB~111117A does not belong to the hostless
class of GRBs \citep{Berger2010a} and because the host exhibits emission lines
this is indicative of a population of relatively young stars. As the majority of
sGRBs \citep{Fong2013b}, the host of GRB~111117A is therefore a late-type galaxy
and is entirely consistent in terms of stellar mass and stellar age with the
general population of sGRB hosts ($\left\langle M _* \right\rangle = 10^{10.1}
M_{\odot}$ and $\left\langle \tau _* \right\rangle = 0.3 $~Gyr)
\citep{Leibler2010}. Being a late-type host, both the stellar mass and sSFR are
entirely within the range expected for the hosts of sGRBs \citep{Behroozi2014}.
Our constraints on the dynamical mass is also well accommodated by the expected sGRB
host halo mass \citep{Behroozi2014}.

@@

A single sGRB at high redshift does little in terms of constraining the redshift
distribution of sGRBs. In particular, other sGRB hosts could be missed because
they are intrinsically fainter and thus the high redshift of GRB~111117A is only
measured due to the brightness of its host. \citet{Berger2014} compiled a
sample of sGRB host luminosities, normalized by the characteristic galaxy
luminosity at their respective redshift, $L_B/L^{\star}_{B}$. To convert the
SED-inferred $M_B$ of GRB~111117A to $L_B/L^{\star}_{B}$, we use the
characteristic absolute $B$-band magnitude of the Schechter function for blue
galaxies ($U - V < 0.25$) in the redshift window $2.0 \leq z \leq 2.5$ from
\citet{Marchesini2007} and obtain $L_B/L^{\star}_{B} = 1.2$.

Using the complete, flux-limited selection of bursts from \citet{DAvanzo2014a},
excluding GRB~111117A and the likely non-sGRB GRB~090426, we have a
statistically homogeneous sample from which we can address the implications of
the redshift of GRB~111117A. Out of the 14 hosts in the sample, 10 (71 per cent)
have measured redshifts and $L_B/L^{\star}_{B}$. Of the complete sample, the
host of GRB~111117A  is brighter than 80 per cent of the hosts with measured
$L_B/L^{\star}_{B}$. Even if we conservatively assume that \textit{all} the
hosts missing $L_B/L^{\star}_{B}$ are brighter than the host of GRB~111117A, the
host is still brighter than > 60 per cent of sGRB hosts. For all 26 hosts with
$L_B/L^{\star}_{B}$ from \citet{Berger2014}, the host of GRB~111117A is brighter
than 73 per cent.

If we assume that we are able to obtain emission-line redshifts from hosts, 0.5
mags fainter \citep[$R < 24.5$~mag;][]{Kruhler2012}, then we would have missed 60
per cent of the redshifts (6 out of 10 hosts), were they at the redshift of
GRB~111117A due to the hosts being too faint. The corresponding number is around
45 per cent (12 out of 26) from the full sample of \citet{Berger2014},
reflecting the lower mean $L_B/L^{\star}_{B}$ of the complete sample. Because
the average SFR of galaxies hosting lGRBs is higher than for galaxies hosting
sGRBs, the fraction of missed burst redshifts is likely higher although the
cosmic SFR evolution could play a role in improving redshift determinability.

A fraction of the bursts missing redshift are host-less but appear to be
spatially correlated with galaxies that are likely at moderate redshifts
\citep{Tunnicliffe2014}, but should some of the remainder be at high redshift,
the missed fraction will increase. If we assume that \textit{all} the bursts
that are missing redshifts are at high-$z$ and missed due to host faintness, 10
out of 14 hosts in the complete sample (71 per cent) would be missed at $z =
2.211$. This serves as an upper limit on the fraction of missed burst redshifts
at high-$z$. Conversely, if all bursts missing redshift are at low redshift and
missed for other reasons, 6 out of 14 hosts (43 per cent) would be missed at $z
= 2.211$. The two limits indicate that we would miss between 43 and 71 per cent
of sGRB hosts at $z \sim 2$ due to host faintness.

@@

Using the new distance, the X-ray derived $N_H$ towards GRB~111117A is the
highest within a complete sample of sGRB hosts and is consistent with the
$N_H-z$ evolution traced by the hosts of lGRBs. The SFR of the host is in the
upper end of the sGRB host SFR distribution and no dust is present. The high
$N_H$ is difficult to reconcile with the large projected host offset and the
absence of dust. One possible explanation could be, that GRB~111117A is formed
through a prompt channel of sGRB formation and originates in a star forming
region located in the outskirts of the host.

@@

Using the age of the universe at the time of explosion allows us to set
constraints on the maximal separation between the engine constituents at the
time of formation. We find that the maximal separation of two NSs at system
formation time is $a_0 < 3.2~R_\odot$, which excludes some of the formation
channels with the longest timescales.

@@

\begin{acknowledgements}
We thank Jens Hjorth and Lise Christensen for useful discussions regarding the interpretation of this event. 
%	
TK acknowledges support through the Sofja Kovalevskaja Award to P. Schady.
% 
JJ acknowledges support from NOVA and NWO-FAPESP grant for advanced
instrumentation in astronomy.
%
NRT and KW acknowledge support from STFC Consolidated
Grant ST/N000757/1
%
CT acknowledges support from a Spanish National Research Grant of Excellence
under project AYA 2014-58381-P and funding associated to a Ramón y Cajál
fellowship under grant number RyC-2012-09984.
%
AdUP acknowledges support from a Ramón y Cajal fellowship, a BBVA Foundation
Grant for Researchers and Cultural Creators,and the Spanish Ministry of Economy
and Competitiveness through project AYA2014-58381-P. Partly based on
observations made with the Gran Telescopio Canarias (GTC).
%
ZC acknowledges support from the Spanish research project AYA 2014-58381-P and
support from Juan de la Cierva Incorporaci\'on fellowships IJCI-2014-21669.
%
RSR acknowledges AdUP's BBVA Foundation Grant for Researchers and Cultural
Creators and support from ASI (Italian Space Agency) through the Contract n. 2015-046-R.0 and from European Union Horizon 2020 Programme under the AHEAD project (grant agreement n. 654215).
%
This research made use of Astropy, a community-developed core Python package for Astronomy \citep{TheAstropyCollaboration2013}. The analysis and plotting has been achieved using the Python-based packages Matplotlib \citep{Hunter2007}, Numpy and Scipy \citep{VanderWalt2011} along with other community-developed packages.
%
This work made use of observations obtained with the Italian 3.6 m Telescopio Nazionale Galileo (TNG) operated on the island of La Palma by the Fundaci\'on Galileo Galilei of the INAF (Istituto Nazionale di Astrofisica) at the Spanish Observatorio del Roque de los Muchachos of the Instituto de Astrof\'isica de Canarias.
%
Based on data from the GTC Archive at CAB (INTA-CSIC) and on observations obtained at the Gemini Observatory, which is operated by the Association of Universities for Research in Astronomy, Inc., under a cooperative agreement with the NSF on behalf of the Gemini partnership: the National Science Foundation (United States), the National Research Council (Canada), CONICYT (Chile), Ministerio de Ciencia, Tecnología e Innovación Productiva (Argentina), and Ministério da Ciência, Tecnologia e Inovação (Brazil).
